---
title: "Tutorial"
author: "Brett Bernhardt Dombrowski"
date: "May 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Data Science Tutorial

##Introduction 

Are different types of coral reefs home to more species of fish than other types of coral reefs? How does water temperature impact the coral reef and the ecosystem that thrives off this living structure? Using two different data sets, Coral Reef Water Temperature Monitoring Protocol Data and National Coral Reef Monitoring Program: Assessment of coral reef fish communities in the U.S. Virgin Islands data, we can answer these questions and discover more findings. This tutorial will walk you through the entire data science pipeline. The first step is to identify the problem or question you want to explore, which we have described above for our tutorial. Next you want to find resources that have data tables with enough entities and key attributes to create informative findings on the subject, which we have also listed above. Then we will show you how to collect and clean the data to make it readable for the machine and user with data curation, parsing, and management tools. Once that data is manageable we will show you how to complete an exploratory data analysis by creating new data tables and graphs to create hypothesis about the data. Then we will show you how to test the hypothesis and use machine learning to provide an analysis for your findings. Finally, we will show you how to curate a message or messages covering insights learned during the tutorial.

#Required Set-up

You will need [R](https://cran.r-project.org/index.html) and the following packages to follow along with the tutorial:
    1. tidyverse
    2. rvest
    3. dplyr
    *If you are using RStudio you can install by typing (install.packages("name_of_package_from_list_above")) in your console.  

We also recommend downloading and using [RStudio](https://www.rstudio.com/products/rstudio/download/) as the development environment to interact with R, this powerful application makes data analysis easier. RStudio also allows for libraries to easily be downloaded in the interactive console. Once in RStudio create a new project by following these steps: File -> New Project -> New Directory -> New Project -> (under project name, type) Tutorial -> Create Project. Then RStudio will redirect you to a blank window in the new project directory. Then you will need to create a new RMarkdown document, to do this follow these steps: File -> New File -> R Markdown -> (under Title, type) tutorial -> Ok. This file allows you to type paragraphs as well as code. To differentiate between code and written prose type this format for entering code: 
```{r any_name_can_be_here_but_cant_repeat_names}

```

Finally, you will need both of the data sets talked about above. Coral Reef Water Temperature Monitoring Protocol Data can be obtained at https://www.nps.gov/im/sfcn/coral-reef-water-temp.htm. National Coral Reef Monitoring Program: Assessment of coral reef fish communities in the U.S. Virgin Islands data can be obtained at https://catalog.data.gov/dataset/national-coral-reef-monitoring-program-assessment-of-coral-reef-fish-communities-in-the-u-s-vir.

#Collect Data

There are multiple methods of retrieving data from a website based on how the data is published. We will be walking through two different methods of data collection. 

The first data set we will be downloading into RStudio is a CSV (comma-separated values) file, which makes the process a bit quicker and simpler. First download the zip file from [Coral Reef Water Temperature Monitoring Protocol Data](https://irma.nps.gov/DataStore/Reference/Profile/2257123). Once this file has downloaded, unzip the file. There should be a folder with 78 CSV Documents. We only want the data sets relating to the U.S. Virgin Islands, so we will highlight only the CSV Documents starting with "VIIS", there should be 12. Open a File Explorer/Finder window and search for the project folder containing the RMarkdown file, Tutorial. Then drag the selected files into the folder, which should add the data sets into the RStudio project. This can be accessed in code by:
```{r example}
data_temp1 <- read_csv("VIIS_HA_1_20181025.csv")
```

The second data set we will be downloading into Rstudio we will scrape from HTML. [National Coral Reef Monitoring Program: Assessment of coral reef fish communities in the U.S. Virgin Islands data](https://ecowatch.ncddc.noaa.gov/erddap/tabledap/CRCP_Reef_Fish_Surveys_USVI.htmlTable?time%2Clatitude%2Clongitude%2Cprimary_sample_unit%2Cstation_nr%2Csample_depth%2Cunderwater_visibility%2Cgrid_id%2Chabitat_cd%2Chabitat_type%2Cdepth_strat%2Cdepth_strat_description%2Csub_region%2Csub_region_name_description%2Cadmin%2Cadministration_description%2Cspecies_nr%2Cspecies_cd%2Cscientific_name%2Ccommon_name%2Clen%2Cnum%2Ctime_seen%2Cprot%2Cstrat%2Cstrat_description%2Cregion%2Cregion_description%2Caccession_url&time%3C=2019-05-20T18%3A20%3A56Z) will bring you to the HTML data set. To scrape this data we will use a CSS selector to get the table. In the inspect element on the website we were able to see that we can select by class to get the data set. Below is the code to scrape the data (I will also include the libraries that are needed to complete the code, including the libraries only needs to be coded once in a document before it is used):


#Clean the Data

Once we have gotten the data from our two sources, we need to make sure that its in proper shape for us to use it. 


As you know from above, the source of our temperature data has many csv files all in the same format, so we will perform the same operations on all of them. We wThe temperature data came to us in a more roundabout way, but the only real thing that we need to do is make sure that none of the entities have blank attributes or "NA". This can be accomplished with dplyr's filter() function that puts a constraint on the entities we want to consider, in this case it only includes entities where none of the fields are NA or blank.
```{r temp-prep}
  clean_temp_NF_1 <- VIIS_NF_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_NF_2 <- VIIS_NF_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")
  
  clean_temp_MB_1 <- VIIS_MB_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_MB_2 <- VIIS_MB_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")

  clean_temp_TK_1 <- VIIS_TK_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_TK_2 <- VIIS_TK_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")

  
  clean_temp_HA_1 <- VIIS_HA_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_HA_2 <- VIIS_HA_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")

  clean_temp_WS_1 <- VIIS_WS_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_WS_2 <- VIIS_WS_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")

  clean_temp_YZ_1 <- VIIS_YZ_1_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")


  clean_temp_YZ_2 <- VIIS_YZ_2_20181025 %>%
  filter(Date_Time != "NA", Temp_C != "NA", Quality != "NA", Interpolation!= "NA", Approval != "NA")

```


With our fish data there is a glaring problem almost immediately, there is an entity for non-sightings of fish. The csv has an entity for each time/place-species combination, most of these entities do not consist of fish-sightings. So what we need to do is remove all entities where the num is 0, because in these instances there are no fish to consider. We can accomplish this with a another simple "filter()" statement below, its just that this time we are constraining our set to be   

```{r fish-prep}
clean_fish_data <- CRCP_Reef_Fish_Surveys_USVI_4490_3467_8907 %>%
  filter(num > 0)

```



##ANALYSIS

We've got our data nice and tidy and are finally ready to start fiddling around with it. We've already introduced you to filter() and %>%, but there are a load more dplyr functions that can play a large role in our analysis of the data.

This following block of code will 
  *use group_by to group by the strat_description
  *use mutate to create a new attribute, that measures the species diversity present at each distinct strat
    *To do this I use n_distinct(), another dplyr function that counts the unique elements
  *use select to cut out the attributes we have no interest in
  *use summarize to change the structure of our table to have each entity be a strat_description
    *we also use this summarize to construct our new table with all sorts of information about the different strat(i.e. amount of fish if laid front to back, and num_sightings)
```{r fish analysis}


mod_data <- clean_fish_data %>%
  group_by(strat_description) %>%
  mutate(fish_diversity = n_distinct(species_nr)) %>%
  select(latitude, longitude, strat_description, fish_diversity, habitat_type, num, len, sub_region_name_description) %>%
  summarize( num_sightings = n(), fish_diversity = mean(fish_diversity), fish_volume = sum(as.numeric(num) * as.numeric(len)))
  
mod_data
```

```{r }


```
